---
title: "Getting started: a start-to-finish dose-response walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: a start-to-finish dose-response walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

This vignette walks the Ward rotavirus human-challenge data from raw counts to a reported risk number, interpreting every output along the way. It answers two questions an end user actually has: *what does my data need to look like going in*, and *what are the results saying coming out*.

```{r setup}
library(singlehit)
```

## Load

The analysis needs **three columns, one row per dose group**: the administered `dose`, and the numbers of subjects who did (`positive`) and did not (`negative`) respond. The package ships this exact shape as `ward_rotavirus`:

```{r load-data}
data(ward_rotavirus)
ward_rotavirus
```

There are two ways in. The **in-memory** path standardizes a data frame you already have:

```{r as-dose-response}
ward <- as_dose_response(ward_rotavirus)
ward
```

The **from-disk** path reads the same data from the bundled text file — this is what you would use for your own delimited file:

```{r read-from-disk}
ward_path <- system.file("extdata", "Ward_rotavirus.txt", package = "singlehit")
ward_from_file <- read_dose_response(ward_path)

identical(ward, ward_from_file)
```

Both routes produce the standardized `dose · positive · negative · total · response` tibble, where `total = positive + negative` and `response = positive / total`. Column names are auto-detected case- and punctuation-insensitively (`pos`, `positive_response`, etc.); override detection with `as_dose_response(data, dose =, positive =, negative =)`. Rows sharing a dose are summed. Fitting requires **at least 3 distinct doses** and **more than 1 total positive response** — non-conforming data is rejected with an explanatory error.

## Keep experiment details

A plain three-column data frame remains valid. You can also add `study_id`,
`host`, `dose_unit`, and `endpoint`. Each supplied identity column must have one
non-blank value across the table. `study_id` means an individual experiment;
`endpoint` describes what was counted, while numeric `response` remains the
observed proportion. Split experiments with different identities into separate
tables before coercion.

```{r metadata-example}
example_path <- system.file("extdata", "dose-response-example.csv", package = "singlehit")
example_trial <- read_dose_response(example_path)
example_trial
```

These example counts are illustrative. For your own experiment, find the blank
CSV template with `system.file("extdata", "dose-response-template.csv", package = "singlehit")`,
copy it into your working folder, and fill it in with a spreadsheet editor.
Repeat experiment details on every row, enter counts and doses as numbers, and
save as CSV. Read it with `read_dose_response("my_trial.csv")`.

Additional metadata columns are preserved. Duplicate doses are combined only
when their metadata agrees; conflicting values produce an error naming the
column. `total` and `response` are recalculated from the counts.

## Run

`analyze_dose_response()` does everything in one call: trend test, model fits, diagnostics, comparison, assessment, and bootstrap uncertainty. A fixed `seed` makes the bootstrap reproducible.

```{r analyze}
analysis <- analyze_dose_response(
  ward,
  bootstrap_times = 1000, # reduced for a fast vignette build; the default is 10000
  resample = "observed",
  seed = 2026
)

analysis
```

The rest of this vignette unpacks each component of that `qdr_analysis` object.

## Interpret each output

### Trend — is there a dose-response signal at all?

```{r trend}
analysis$trend
```

`dose_trend_test()` runs a one-sided log-dose trend test. `passes = TRUE` means a **monotonic increasing** dose-response relationship was detected — response rises with dose.

> **What this is really saying:** if `passes` were `FALSE`, there is no upward dose-response signal to model, and fitting mechanistic curves would not be warranted. Here it passes, so we proceed.

### Goodness of fit — does each model fit the data at all?

```{r goodness}
analysis$goodness_of_fit
```

This is the **absolute** check. Each model's `deviance` is compared against a chi-squared `critical_value`; `good_fit` is `TRUE` when the deviance falls below it (equivalently, `p_value` \> 0.05). `assessment` records `adequate` or `inadequate`.

> **What this is really saying:** does this model describe the data acceptably on its own terms? A model can "win" a comparison and still fit badly — this is the gate that catches that.

### Comparison — which model wins, and did the extra parameter earn its keep?

```{r comparison}
analysis$comparison
```

This is the **relative** check. The two-parameter Beta-Poisson is tested against the one-parameter Exponential (its nested limiting case) via the `deviance_difference` on the extra degree of freedom, alongside `AIC`/`BIC` and their deltas. `significant_improvement` and `preferred` flag the winner.

> **What this is really saying:** the Beta-Poisson always fits at least as well because it has an extra parameter — `significant_improvement` asks whether that extra flexibility bought a *statistically real* improvement, or just curve-fit noise.

### Assessment — the headline verdict

```{r assessment}
analysis$assessment[, c("model", "appropriate", "preferred", "recommendation", "conclusion")]
```

`build_model_assessment()` fuses the absolute and relative checks into one `recommendation` label per model:

-   `recommended` — fits adequately **and** is preferred.
-   `acceptable_alternative` — fits adequately but is not preferred.
-   `preferred_but_inadequate` — preferred but does **not** fit adequately (a red flag: the "best" model is still wrong).
-   `not_recommended` — neither.

> **What this is really saying:** this is the one row a risk assessor reads first. `recommended` means "use this model with confidence"; the `conclusion` column spells it out in plain language.

### Consensus — an experimental cross-check

```{r consensus}
consensus_model_decision(analysis)
```

`consensus_model_decision()` treats the chi-squared, AIC, and BIC selections as three votes and reports the `agreement` level (`unanimous`, `majority`, or `no_consensus`).

> **What this is really saying:** an experimental sanity check on the relative selection. It does **not** replace the absolute goodness-of-fit gate above — a unanimously-voted model can still fit inadequately.

### Bootstrap intervals — the final risk numbers and their uncertainty

```{r bootstrap-ci}
bootstrap_confint(analysis$bootstraps$beta_poisson, levels = c(0.95))
```

The bootstrap refits the model to resampled data many times and reports percentile confidence intervals — not just for the parameters (`alpha`, `n50`) but for the **effective doses** `ed10` and `ed50`: the doses producing a 10% and 50% response.

> **What this is really saying:** these are the deliverable. `ed50` is the median infectious dose; its interval is the honest uncertainty around it, driven by how much data you have and how noisy it is.

## Plots

`autoplot()` on the analysis shows each fitted curve over the observed points:

```{r plot-analysis}
ggplot2::autoplot(analysis)
```

`autoplot()` on a bootstrap object shows the sampling distribution of the fitted parameters — the visual form of the intervals above:

```{r plot-bootstrap}
ggplot2::autoplot(analysis$bootstraps$beta_poisson)
```

## The final answer

Read the recommended model's `ed50` and its interval, and you have the number a risk assessor reports:

```{r final-answer}
# Which model did the assessment recommend?
recommended <- analysis$assessment$model[analysis$assessment$recommendation == "recommended"][1]
recommended

# Its bootstrap intervals, then keep only the ed50 row.
ci <- bootstrap_confint(analysis$bootstraps[[recommended]], levels = c(0.95))
ci[ci$term == "ed50", ]
```

That single row — the median infectious dose of the recommended model with its 95% percentile interval — is the deliverable: *the dose at which half of exposed subjects are expected to be infected, with a defensible bound on how sure we are.*

## Which experiments can be pooled?

Use the bundled synthetic collection to try pooling before using your own
studies. It contains three artificial experiments, A, B, and C, with the same
host, dose unit, and infection endpoint. Each experiment has four doses and
100 subjects per dose. A and B have similar curves; C has a lower response.
These are teaching data, not observations from a real pathogen.

### Read and split the collection

```{r pooling-import}
pooling_path <- system.file("extdata", "pooling-example.csv", package = "singlehit")
raw_trials <- readr::read_csv(pooling_path, show_col_types = FALSE)
collection <- list(
  pathogen = "Synthetic example",
  datasets = split(raw_trials, raw_trials$study_id)
)
names(collection$datasets)
```

The output names A, B, and C. Keep the pathogen name once on the collection.
Split by `study_id` before coercion because it identifies an individual
experiment. `read_dose_response()` handles one experiment at a time and rejects
mixed identity values. Check that hosts, units, and endpoints match before
pooling; the current pooling functions do not enforce that comparison.

### Inspect every combination

```{r pooling-combinations}
report <- poolability_combinations(collection$datasets)
report[c("combination", "model", "p_value", "poolable", "converged")]
```

Each row tests one combination under one model. At the default `alpha = 0.05`,
A+B passes under both models, with p-values around 0.94 and 0.93. Every
combination containing C fails. All fits converged, so this example has no
optimizer failure flags.

`poolable = TRUE` means the test did not detect a significantly worse fit when
sharing one parameter set. It does not prove equivalence or adequate absolute
model fit. A passing pair can fail after another experiment is added, so read
the row for the complete combination you plan to pool. These are exploratory
p-values without multiple-testing adjustment.

### Select non-overlapping groups

```{r pooling-groups}
groups <- group_datasets(collection$datasets, method = "exhaustive")
groups
```

A and B share group 1; C is in group 2 for both models. Group numbers are labels,
not rankings. The exhaustive method considers all divisions into groups and
returns one selected grouping per model. The default agglomerative method
instead merges compatible groups greedily. Exhaustive grouping and combination
reporting both default to a maximum of six experiments.

You can also inspect just the proposed A+B pool:

```{r pooling-selected-pair}
pair <- poolability_test(collection$datasets[c("A", "B")])
pair[c("model", "p_value", "poolable", "converged", "conclusion")]
```

For this example, A and B are candidates for a shared fit; keep C separate.
The report and grouping functions test poolability and return assignments.
They do not create a pooled analysis or establish goodness of fit. A later
pooled fit must preserve separate trial observations at repeated doses;
simply coercing stacked trials would aggregate those observations or reject
mixed identities.
